---
title: "Beer"
author: "Jacob Gipson and Tavin Weeda"
date: "10/9/2021"
output: html_document
---

# Beer EDA
```{r}
#upload beer and breweries data
beers = read.csv("C:\\Users\\tavin\\OneDrive\\Desktop\\DDS\\Project 1\\Beers.csv", header = TRUE)
breweries = read.csv("C:\\Users\\tavin\\OneDrive\\Desktop\\DDS\\Project 1\\Breweries.csv", header = TRUE)
```

### Question 1: How many breweries are in each state?
```{r}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(RColorBrewer)
####This creates a custom color palette

  # Classic palette BuPu, with 4 colors
  coul <- brewer.pal(4, "PuOr") 

  # Add more colors to this palette :
  coul <- colorRampPalette(coul)(51)

##Make Brewery count graph
 
brewerycount = data.frame(table(breweries['State']))
colnames(brewerycount) = c('State', 'BrewCount')
brewerycount
brewerycount %>%
  ggplot(aes(x = reorder(State, BrewCount), y = BrewCount, fill = State)) +
  theme_classic() +
  geom_bar(stat = "identity",fill=coul)+
  coord_flip() + 
  theme(legend.position = "none") + 
  ylab("Number of Breweries") +
  xlab("State") +
  ggtitle("Number of Breweries in Each State") 
  
```
  
### Question 2: Merge beer data with the breweries data. Print the first 6 observations and the last six observations to check the merged file.  (RMD only, this does not need to be included in the presentation or the deck.)
```{r}
colnames(breweries)[1] = "Brewery_id"
colnames(breweries)[2] = "Brewery Name"
fullbeer <- left_join(beers, breweries, by = "Brewery_id")
head(fullbeer, n = 6)
tail(fullbeer, n = 6)
```

### Question 3: Address the missing values in each column.
## We filled in missing ABV and IBU values with averages for the beer's style. 
```{r}
library(tidyverse)
#Number of missing ABV values
sum(is.na(fullbeer$ABV))
#Number of missing IBU values
sum(is.na(fullbeer$IBU))

fullbeer$Style = as.factor(fullbeer$Style)
styles_list<-unique(fullbeer$Style)

#Fill in missing IBU Values with mean IBU value of the style of beer
 for(sty in 1:length(styles_list)){
    l<-fullbeer[fullbeer$Style==styles_list[sty],]
    m<-l[!(is.na(l$IBU)),]
    n<-mean(m$IBU)
   fullbeer$IBU[fullbeer$Style==styles_list[sty] & is.na(fullbeer$IBU)]<-n
 }

#Fill in missing ABV Values with mean ABV value of the style of beer
 for(sty in 1:length(styles_list)){
    l<-fullbeer[fullbeer$Style==styles_list[sty],]
    m<-l[!(is.na(l$ABV)),]
    n<-mean(m$ABV)
   fullbeer$ABV[fullbeer$Style==styles_list[sty] & is.na(fullbeer$ABV)]<-n
 }

#No more missing ABV values
sum(is.na(fullbeer$ABV))
#We still have 52 missing IBU values
sum(is.na(fullbeer$IBU))


#We assume all mead and ciders should have an IBU of 0 as they generally do not come with an IBU value.
#The rest of the styles were looked up.
fullbeer$IBU[fullbeer$Style == "Cider"] = 0
fullbeer$IBU[fullbeer$Style == "Mead"] = 0
fullbeer$IBU[fullbeer$Style == "Rauchbier"] = 0
fullbeer$IBU[fullbeer$Style == "Shandy"] = 0
fullbeer$IBU[fullbeer$Style == "American Malt Liquor"] = 0
fullbeer$IBU[fullbeer$Style == "Kristalweizen"] = 40
fullbeer$IBU[fullbeer$Style == "Flanders Red Ale"] = 59
fullbeer$IBU[fullbeer$Style == "Braggot"] = 12
fullbeer$IBU[fullbeer$Style == "Low Alcohol Beer"] = 0

#No more missing IBU values
sum(is.na(fullbeer$IBU))

# 5 rows are missing styles.  2527 and 1635 are actual beers.  The other 3 are not specific beers so they were removed
fullbeer$Style = as.character(fullbeer$Style)
fullbeer$Style[fullbeer$Beer_ID == 2527] = "Märzen / Oktoberfest"
fullbeer$Style[fullbeer$Beer_ID == 1635] = "Scottish Ale"


fullbeer$Style = as.factor(fullbeer$Style)

#Now the two beers have a style
fullbeer[fullbeer$Beer_ID == 67,]
fullbeer[fullbeer$Beer_ID == 1635,]
fullbeer<-fullbeer[!(fullbeer$Beer_ID == 2210 | fullbeer$Beer_ID==1796|fullbeer$Beer_ID==1790),]
```

### Question 4: Compute the median alcohol content and international bitterness unit for each state. Plot a bar chart to compare.
```{r}
library(dplyr)
fullbeer$State = as.factor(fullbeer$State)
medians = fullbeer %>%
  group_by(State) %>%
  summarise(medianABV = median(ABV), medianIBU = median(IBU))

medians$medianABV<-round(medians$medianABV*100,digits=1)

medians %>%
  ggplot(aes(y = reorder(State, medianABV), x = medianABV, fill = State)) +
  geom_bar(stat = "identity",fill=coul) +
  theme_clean() +
    
  theme(legend.position = "none") + 
  ylab("State") +
  xlab("ABV (%)") +
  ggtitle("Median Beer ABV in Each State")+coord_cartesian(xlim=c(3,7))

medians %>%
  ggplot(aes(y = reorder(State, medianIBU), x = medianIBU, fill = State)) +
  geom_bar(stat = "identity",fill=coul) +
  theme_clean() +
  theme(legend.position = "none") + 
  ylab("IBU") +
  xlab("State") +
  ggtitle("Median Beer IBU in Each State")+coord_cartesian(xlim=c(15,70))

```

### Question 5: Which state has the maximum alcoholic (ABV) beer? Which state has the most bitter (IBU) beer?
```{r}
#Highest ABV
fullbeer[which.max(fullbeer$ABV),]
#Most Bitter - Highest IBU
fullbeer[which.max(fullbeer$IBU),]
```

### Question 6:	Comment on the summary statistics and distribution of the ABV variable.
```{r}
summary(fullbeer$ABV)
fullbeer$ABV<-round(fullbeer$ABV*100,digits=1)

fullbeer %>%
  ggplot(aes(x = ABV, fill = "red"))+
  geom_histogram() +
  theme_clean() + 
  theme(legend.position = "none") +
  ylab("Beer Count") +
  ggtitle("Histogram of Beer ABV")+
  xlab("ABV (%)")

fullbeer %>% 
  ggplot(aes(x = ABV, color = "red")) +
  geom_density() + 
  theme_clean() +
  theme(legend.position = "none") +
  ggtitle("Density plot of Beer ABV")

# Very right skewed distribution. Most beers are in the range of 0.04 - 0.06.
```

### Question 7: Is there an apparent relationship between the bitterness of the beer and its alcoholic content? Draw a scatter plot.  Make your best judgment of a relationship and EXPLAIN your answer.
```{r}
library(ggpmisc)
my.formula = y ~ x
fullbeer %>%
  ggplot(aes(x=ABV,y=IBU))+
  geom_point(position = "jitter") +
  geom_smooth(method="lm", color = "red") +
  stat_poly_eq(formula = my.formula, 
               aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
               parse = TRUE,
               label.y = 0.87,
               eq.x.rhs = "ABV") +
  theme_clean() +
  ggtitle("IBU vs ABV")
```
### Question 8: Use KNN classification to investigate the difference between IPAs and other typles of Ales with respect to IBU and ABV.
```{r}
library(class)
library(caret)

#First we pull all the beers that are IPA's or have Ale in the name
IPAsandAles<-fullbeer[grepl("IPA|Ale",fullbeer$Style),]

#Then we split the data into 2 groups: IPA's and Ales
IPAs <- IPAsandAles[grepl("IPA|India Pale Ale", IPAsandAles$Style),]
Ales <- IPAsandAles[!grepl("IPA|India Pale Ale", IPAsandAles$Style),]

#Create a new column to identify each beer as IPA or Ale and then re-combine datasets
IPAs$Class = "IPA"
Ales$Class = "Ale"
IPAsandAles <- rbind(IPAs, Ales)

#Create new data frame from IPAsandAles that only includes IBU, ABV, and Class information.
IPAAle <- data.frame(ABV = IPAsandAles$ABV, 
                     IBU = IPAsandAles$IBU, 
                     ZABV = scale(IPAsandAles$ABV),
                     ZIBU = scale(IPAsandAles$IBU),
                     Class = IPAsandAles$Class)

#We are first splitting our data into a training set and a validation set.

#Train 80% validate 20%
set.seed(1)
trainIndices = sample(1:dim(IPAAle)[1],round(.8* dim(IPAAle)[1]))
train = IPAAle[trainIndices,]
validate = IPAAle[-trainIndices,]

#To tune the hyperparameters, we did a  cross validation and a for loop using train/test splits.  

iterations = 500
numks = 30
masterAcc = matrix(nrow = 1, ncol = numks)
masterSens = matrix(nrow = 1, ncol = numks)
masterSpec = matrix(nrow = 1, ncol = numks)

  for(i in 1:numks)
  {
    classifications = knn.cv(train[,c(3,4)],train$Class, prob = TRUE, k = i)
    table(classifications,train$Class)
    CM = confusionMatrix(table(classifications,train$Class))
    masterAcc[1,i] = CM$overall[1]
    masterSens[1,i]=CM$byClass[1]
    masterSpec[1,i]=CM$byClass[2]
  }

MeanAcc = colMeans(masterAcc)
MeanSens=colMeans(masterSens)
MeanSpec = colMeans(masterSpec)

#Make plots to see the effect of K
par(mfrow=c(1,3))
CVAcc = data.frame(K = seq(1,numks,1), MeanAcc = MeanAcc)
CVAcc %>% 
  ggplot(aes(K, MeanAcc)) +
  geom_point() +
  geom_line() +
  theme_clean() + 
  ylab("Accuracy")+
  ggtitle("Effect of K on Accuracy (CV)")

CVSens = data.frame(K = seq(1,numks,1), MeanSens = MeanSens)
CVSens %>% 
  ggplot(aes(K, MeanSens)) +
  geom_point() +
  geom_line() +
  theme_clean() + 
  ylab("Sensitivity")+
  ggtitle("Effect of K on Sensitivity (CV)")

CVSpec = data.frame(K = seq(1,numks,1), MeanSpec = MeanSpec)
CVSpec %>% 
  ggplot(aes(K, MeanSpec)) +
  geom_point() +
  geom_line() +
  theme_clean() + 
  ylab("Specificity")+
  ggtitle("Effect of K on Specificity (CV)")



# This for loop creates 500 iterations of train/test splits (from the training data) to tune k
iterations = 500
numks = 30
masterAcc = matrix(nrow = iterations, ncol = numks)
masterSens = matrix(nrow = iterations, ncol = numks)
masterSpec = matrix(nrow = iterations, ncol = numks)
splitPerc=.75

for(j in 1:iterations)
{
  trainIndices = sample(1:dim(train)[1],round(splitPerc * dim(train)[1]))
  train1 = train[trainIndices,]
  test1 = train[-trainIndices,]
  for(i in 1:numks)
  {
    classifications = knn(train1[,c(3,4)],test1[,c(3,4)],train1$Class, prob = TRUE, k = i)
    CM = confusionMatrix(table(classifications,test1$Class))
    masterAcc[j,i] = CM$overall[1]
    masterSens[j,i] = CM$byClass[1]
    masterSpec[j,i] = CM$byClass[2]
  }
}

MeanAcc = colMeans(masterAcc)
MeanSens = colMeans(masterSens)
MeanSpec = colMeans(masterSpec)

kAcc = data.frame(K = seq(1,numks,1), Accuracy = MeanAcc)
kAcc %>% 
  ggplot(aes(K, Accuracy)) +
  geom_point() +
  geom_line() +
  theme_clean() + 
  ylab("Accuracy")+
  ggtitle("Effect of K on Accuracy (Train/Test Split)")

kSens = data.frame(K = seq(1,numks,1), Sensitivity = MeanSens)
kSens %>% 
  ggplot(aes(K, Sensitivity)) +
  geom_point() +
  geom_line() +
  theme_clean() + 
  ylab("Sensitivity")+
  ggtitle("Effect of K on Sensitivity (Train/Test Split)")

kSpec = data.frame(K = seq(1,numks,1), Specificity = MeanSpec)
kSpec %>% 
  ggplot(aes(K, MeanSpec)) +
  geom_point() +
  geom_line() +
  theme_clean() + 
  ylab("Specificity")+
  ggtitle("Effect of K on Specificity (Train/Test Split)")

#We settled on k=25 as it balanced Accuracy, Sensitivity and Specificity the best.
classifications = knn(train[,c(3,4)],validate[,c(3,4)],train$Class, prob = TRUE, k = 19)
CM = confusionMatrix(table(classifications,validate$Class))
confusionMatrix(table(classifications,validate$Class))

```


###Exploring a Naive Bayes model
```{r}
##Naive Bayes Exploration

library(e1071)
model = naiveBayes(Class~ABV+IBU,data = train)
confusionMatrix(table(predict(model,validate[,c(1,2)]),as.factor(validate$Class)))
```

###Other Useful Inference: Most and Least numerous type of beer by State
```{r}
count_by_state<-fullbeer%>%group_by(State)%>%count(Style)
max_counts_by_State<-count_by_state%>%group_by(State)%>%summarize(n=max(n))
style_by_state<-left_join(max_counts_by_State,count_by_state,by=c("State","n"))
style_by_state$Style = as.factor(style_by_state$Style)

summary(style_by_state)
most_popular_styles = 
  style_by_state %>%
  group_by(Style) %>%
  summarise(n = n())

most_popular_styles %>%
  ggplot(aes(x = reorder(Style, n), y = n, fill = Style)) +
  geom_bar(stat = "identity") +
  theme_clean() +
  coord_flip() + 
  theme(legend.position = "none") + 
  ylab("Frequency of most numerous beers in each state") +
  xlab("Style") +
  ggtitle("Most numerous beers in Each State")

IPAState = count_by_state%>%group_by(State)%>%filter(Style == "American IPA")
IPAState %>%
  ggplot(aes(x = reorder(State, n), y = n, fill = State)) +
  geom_bar(stat = "identity") +
  theme_clean() +
  coord_flip() + 
  theme(legend.position = "none") + 
  ylab("Number of American IPA's") +
  xlab("State") +
  ggtitle("Number of American IPA's in each state")

#We could make the conclusion that IPA's could be introduced into those markets that don't have many IPA's as that is the most popular class of beers in the states. 

max_style_by_state = style_by_state %>% group_by(State) %>% summarize(n= max(n))
summary(as.factor(style_by_state$Style))
```